Closing the gap between statistical and scientific workflows for improved forecasts in ecology

### 2025-09-18 revision/r2/forecastflows_r2.tex
working on revisions

We argue that many current debates and diverging forecasts are driven by the incoherent and disconnected workflows used today in ecology \citep{Loreau2022, Talis2023, Johnson2024}. Research estimating biodiversity trends has become focused on methodological aspects because the current workflow fails to examine the gap between ideal and available data, and rarely tests for predictive accuracy that could scale up to allow forecasting. At the same time, process-based models developed for forecasting often evolve through the addition of new separate layers or components because the current workflow rarely examines the model as a functioning whole and thus ignore major problems (e.g., non-identifiability, discussed below). Thus, it encourages new parts that are often disconnected from the original research aim, its data stream, and the previous scientific insights.

@@

\begin{figure}
	\centering
	\includegraphics{../figures/figure_modeldataspaces_revised}
	\caption{Different trajectories of current (b-c) and proposed (d) workflows through model-data space during model development, from pre-model/pre-data to post-model/post-data quadrants.
	Panel A serves as a legend, showing where model development should begin (carefully thinking about the hypotheses to test) and when inference should be made, for either explanatory purpose (estimation) or forecasting (prediction).
	Panel B illustrates that, in current trend estimation approaches, most time is spent in the post-model/post-data quadrant, where models are directly fitted to empirical data---often with no prior evaluation of the model.
	Panel C illustrates the typical process-based model development where a lot of time is spent building and calibrating (fitting) separate submodels (represented by the separated ovals) without clear distinction between model building and model calibration and without evaluating the full model structure a priori. Additionally, the specific hypotheses the model is designed to test are rarely clearly stated.
	Panel D illustrates the workflow highlighted in this paper (see main text and Figure 2 for details). Ideally, it begins in the pre-model/pre-data quadrant, where the research question is defined. The next step moves to the post-model/pre-data quadrant, where a model is built---independently of the data. This is where we should spend a good deal of time evaluating model behavior using simulated data. Once we are satisfied with our model, we can move to the post-model/post-data quadrant, where the model is fitted to real data. Then, we perform retrodictive checks to compare model predictions to observations, which likely give us some feedback to refine our model.}
	\label{fig:modeldata}
\end{figure}

@@

Quantitative science relies on a model-based framework to confront hypothesis with data \citep{Chamberlin:1965cd}. In an idealized scientific method, we would formulate a research question and hypotheses, design an experiment accordingly, build a model, collect data, and use this data to inform our model and differentiate between hypotheses. This method underlies much of the recent pre-registration movement, where hypotheses and methodology are defined prior to data collection \citep{Nosek2018}.
But this idealized method often does not apply to the reality of ecological research. Many important questions cannot be addressed through controlled experiments and replications. In such cases, we must rely on existing, heterogeneous datasets alongside uncertain and incomplete theory to provide a large-scale and long-term perspective \citep{Hilborn1997}. Indeed, most macroecological insights have emerged from exploring patterns in these datasets (exploratory data analysis).

This reality should drive researchers to use more robust and coherent methods. But the current workflows combined with the challenges ecologists are facing---both in terms of data complexity and societal needs---instead may lead to persistent problems.
Trend estimation has focused mostly on fitting a model to empirical data (\llabel{quad1}i.e. post-model/post-data, figure \ref{fig:modeldata}b)---without the checks (\llabel{quad2}post-model/pre-data) and likely feedbacks that often highlight uncertainty and related limitations in the model and/or data (figure \ref{fig:modeldata}b). For forecasting, researchers have focused on making predictions with increasingly complex mechanistic models (figure \ref{fig:modeldata}c), frequently obscuring the steps underlying model building and parameterization (\llabel{quad3}and without a clear distinction between pre- and post-data). Researchers often calibrate the different parts of these models separately, and fix some parameter values based on experiments and expert knowledge, to avoid problems when trying to fit the model as a whole.
Addressing these problems while accounting for the realities of working with ecological data requires a more comprehensive workflow.

We argue a workflow that moves \llabel{quad4}along the data-model space in a coherent sequence of steps (figure \ref{fig:modeldata}d) could reduce many of these problems and thus improve ecological science.
The first step of this workflow is to define an explicit research question and formulate hypotheses (step 1, figure \ref{fig:workflow}). This involves making clear assumptions about the most influential drivers, within the specific context of our study. This should guide the construction of a narrative model of how we believe the system works, focusing on the mechanisms that could generate the data we observe, including the observational error (\llabel{quad5}pre-model/pre-data, figure \ref{fig:modeldata}d).  
\llabel{causalinf} This step of carefully formulating hypotheses has gained increasing attention in the ecological literature \citep{Grace2020}. From this narrative, we can then develop a mathematical model---an ensemble of equations that encapsulates our knowledge and is designed to answer our research question (step 2, figure \ref{fig:workflow}). Generally, starting with a relatively simple model that we could refine later makes understanding the model and how it interacts with empirical data easier \llabel{exampleworkflow1}(see example workflow we provide as a supplementary file). At this stage, prioritizing biologically meaningful parameters is crucial, as it allows us to have a sense of plausible parameter values. This means choosing a model formulation where each parameter corresponds to an interpretable behavior (which sometimes requires considering alternative parameterization).

With a model in place, the next step focuses on testing and understanding it via data simulation (step 3, figure \ref{fig:workflow}). `Fake' or `test' data are generated directly from the model by fixing parameters \llabel{firstsim}to some reasonable values, which is straightforward if the parameters are interpretable, and from `fake' (simulated) predictor data. 
We then fit our model to this simulated dataset and evaluate its ability to recover the prescribed values (\llabel{quad6}post-model/pre-data, figure \ref{fig:modeldata}d). At this stage, the focus is on understanding the model, so we may need to spend time thinking about whether each parameter is realistic or not. This is also a step for making sure the model is working as expected (see example workflow). \llabel{repeatsim} Ideally, this data simulation step should be repeated several times with different parameter values, sampled within a reasonable range. \llabel{priorPC} In a Bayesian framework, simulated parameters can be sampled from the prior distribution (allowing us to check the implications of our priors \citealp{Gelman2020}), while in other frameworks they can be sampled from a range of biologically reasonable values (which is very similar to the concept of a prior).

\begin{figure}[h]
	\centering
	\vspace*{-0.5cm}
    \hspace*{-1.5cm}
	\includegraphics{../figures/figure_worflow_wsteps}
	\caption{The workflow we propose here, which builds from recent advances in workflows \citep[][]{betanworkflow,Gelman2020,Schad2020,grinsztajn2021,vandeschoot2021,Wolkovich2024}, focuses on iterative feedbacks between the research question, model and data. Most research should start at Step 1, with the research question, followed by extensive model building and evaluation using simulated data (Steps 2 and 3) before proceeding to fitting the model on empirical data and examining it through retrodictive checks (Steps 4 and 5). Within this workflow, forecasting is a natural output of the process and not a separate process or one available for only certain modeling approaches.  The integration of observations (the data stream) occurs at Step 4---only after the model has been thoroughly evaluated---and can also highlight opportunities to collect new data and enrich the data stream. This figure is expanded from Figure \ref{fig:modeldata}d.}
	\vspace*{-0cm}
	\label{fig:workflow}
\end{figure}

Once we are confident about our model structure, we can introduce real data as part of an initial model fitting step (step 4, figure \ref{fig:workflow}). This way, we obtain parameter estimates constrained by observations (\llabel{quad7}post-model/post-data, figure \ref{fig:modeldata}d). 
These parameter estimates lead to the second data simulation step, this time using our fitted model parameters to generate predictions (step 5, figure \ref{fig:workflow}). This---which we call a retrodictive check (\citealp{betanworkflow}, also called a \llabel{PPC}posterior predictive check in a Bayesian workflow,  \citealp{Gelman2020})---allows model output to be compared to observations \llabel{suppfig1}(see supplementary figure 1). \llabel{howtoPPC} This is a powerful approach, in part because checks should be tailored to assess key features of the data and model output together in regard to our research question, but this also means there is no standard way to do this \citep{Gelman2020}.
It is only once all steps have been completed that we can interpret parameter values with respect to our research question.
The workflow encourages a focus on the full model, where any parameter (such as a trend estimate) must be carefully interpreted alongside others, as all are fundamental components that shape both inference and forecasting.

Within such a workflow, forecasting emerges as a natural outcome. Rather than being a final goal, forecasting only involves jointly modeling new circumstances along with the original data. This workflow could also bridge gaps in certain areas between exploratory analyses and developing research questions. For example, in macroecological studies model building is often informed by patterns in the data. The adoption of this workflow would make the exploratory stage more transparent---as an explicit preliminary step for entering the workflow---and would compel researchers to more clearly develop a research question before extensive model fitting (figure 1).

A key feature of this workflow is the central role of data simulation, which introduces two feedback loops.
The first feedback arises when we evaluate the model on simulated data.
The failure of the model to recover known parameter values and handle the complexity of the simulated data should prompt reconsidering the model, or even reformulating the research question.
Further, this step might reveal that some parameters are highly non-identifiable (meaning the parameter(s) cannot be uniquely estimated), flagging the need to change the model structure---before incorporating empirical observations. 
The second feedback loop comes from the retrodictive check. 
Discrepancies here may indicate a missing key driver, and suggest the current model is too simplistic. We can refine the model to integrate the missing process(es)---if we can identify them---and return to the start of the workflow. Insights from the retrodictive check can also lead us to introduce additional complexity when simulating fake data, such as phylogenetic structure \llabel{suppfig2}(see supplementary figure 1) or observational biases (e.g. unbalanced data). For example, if a researcher realizes their empirical data is geographically biased, this bias should be built into the model and thus into this data simulation step. This iterative evaluation of the model moves beyond a simple reliance on goodness-of-fit metrics. At each iteration, we are able to evaluate the model behavior, both with simulated and real data, taking into account our expert knowledge of the ecological processes. \llabel{broadcontext}These feedback loops are applied currently in only in Bayesian frameworks (to our knowledge), but could and should extend across other modeling approaches given their power to improve models and, in turn, scientific understanding.

@@

We argue that an improved workflow that required data simulation and retrodictive checks would lead to larger model advances and a greater recognition of uncertainty---thus highlighting likely consistency in estimates across models---that could better aid policy.  Using the workflow would make it more obvious that what now appear as major discrepancies are shifts in point estimates that are generally all in the same uncertainty space \citep{Johnson2024}---and it would challenge modelers to show major predictive advances, which is not currently part of the process. Explanatory power in most models of observational data is usually very low \citep{low2014rising,moller2002much} and thus tests of models' predictions rarely expected. But the workflow highlights that predictions from the model---what we call retrodictive checks---are part of the process of science, and critical to testing for what may be missing in a model \llabel{suppfig3}(see supplementary figure 1).

@@

Applying the workflow to current trend estimates could importantly highlight the best way to improve data collection for more reliable estimates. Returning to the example of a global estimate of trends in vertebrate populations of species over time (see Box) and applying our proposed workflow would mean more efforts to define the goal and question---is it a simple global estimate?
Or a need to also find which species are declining most, including those that may have poor or no data? From there a generative model using simulated data for testing could incorporate many aspects of the populations, and data, that are often only included in `null' or `synthetic data generation' currently \citep{Buschke2021,mcrae2025utility}, but could be built into the models fit to the empirical data. Eventually fitting the empirical data and performing retrodictive checks would likely highlight major missing components of the generative model \llabel{suppfig4}(see supplementary figure 1) and, ultimately, this would help inform our global estimates of mean trends \llabel{exampleworkflow2}(see example workflow we provide as a supplement file). For example, certain populations are recovering for very specific reasons (e.g., elephants in regions where the ivory trade drove declines in the past) that perhaps should be modeled. From this model, what data are most critically needed to address the updated aims would become clearer and could drive new data collection \citep{toszogyova2024mathematical}.

@@

Ecological forecasting is a broad field with a diverse range of methods. Across these methods, process-based modeling is often considered the gold standard for forecasting in ecology \citep{Urban2016, Pilowsky2022} and beyond because it is based on mechanistic understanding that should drive accurate out-of-sample predictions.  
This focus on a mechanistic basis, however, generally drives newer models to incorporate greater complexity---and an ever-growing number of parameters. This can make it difficult to increase scientific understanding \citep{Franklin2020} and suggest or model potential policies. 
In principle, increasing model complexity can be beneficial by enhancing mechanistic realism, better representing more complex dynamics, and \llabel{quant2}reducing uncertainty by restricting the effective parameter space. In practice, however, this is not always the outcome. In process-based modeling for climate forecasts, the uncertainty range on the effect of increasing CO$_{2}$ concentration on temperature have remained largely unchanged over decades of increasing model complexity \citep{Zelinka2020}. This has driven calls for more rigorous and transparent calibration processes \citep{balaji2022general}. Similar concerns arise in ecology, where strong disagreements persist about the effect of climate change on future species distributions \citep{Cheaib2012} and ecosystem dynamics \citep{Lovenduski2017} after decades of modeling work.
These uncertainties have large implications beyond ecology, as they influence simulations of biosphere-atmosphere interactions and, ultimately, future climate projections \citep{Bonan2018, simpson2025confronting}.
Some researchers now advocate for simplifying models, to avoid over-parametrization when the data provide little information to constrain some parameters \citep{Wang2017, Harrison2021}. These arguments stress that the strength of process-based models is their interpretability, and thus their parameters should represent biological processes, and the values of these parameters are of interest.
If a model becomes too complex, however, it may become a black box, \llabel{blackbox}where many parameters are effectively hidden and can take on unreasonable values. 
Further, with so many parameters understanding the sources of uncertainty and how they propagate through the model may become nearly impossible (see Box).
Each additional process and parameter can increase overall uncertainty to the point where model projections lose their usefulness for decision makers \citep{Saltelli2020}.

@@

Applying the workflow to process-based models would help open the black box and thus could identify and fix model problems before forecasting.
Each successive step of model development in the proposed workflow may highlight problems and a path to solutions. Incorporating data simulation would introduce a crucial step between model building and data fitting (also called calibration), ensuring a clear delineation between the two, and help expose potential parameter identifiability issues earlier in the model design. 
Uncovering identifiability issues would likely force researchers to begin with a simpler version of the model, which they could build on iteratively, testing for support---or lack thereof---when adding model complexity. The workflow may also highlight model problems by requiring more explicit model calibration (i.e. data fitting), which is currently hidden within opaque `model building', making it easy to hide non-identifiability. Once found, non-identifiable parts of models could be addressed by reformulating the mathematical structure of certain processes, or finding ways to apply additional constraints (e.g., narrower ranges for certain parameters or developing new hypotheses that target the appropriate level of complexity). The resulting process-based models would likely be simpler and thus more tractable for quantifying parameter uncertainty and propagating uncertainty through to projections.

@@

\centerline{\bf Trend estimation} 
\vspace*{2mm}
\begin{minipage}[t]{\linewidth}
    \includegraphics[width=\linewidth]{../figures/trendestimation_details}

@@

% Trends - outline current problem
% In the current workflow for estimating trends over time, a new model with a new estimate often leads to a paper (see figure A above) because ecologists spend less time interrogating their models with simulated data, or their model performance fit to empirical data.
In the current workflow for estimating trends over time, a model is typically chosen a priori to fit to data, yielding parameter estimates that are treated as true and often translated directly into published findings and conclusions, without interrogating the model using simulated data and carefully evaluating its performance (see example workflow we provide as a supplementary file). Any such analysis must be treated with caution, because different model variants can yield very different estimates (hence different conclusions), and published results may not prove robust when confronted with future reanalysis of the same data.
The Living Planet Index (LPI, \url{www.livingplanetindex.org}), which aims to include long-term data on vertebrate populations of species across the globe, is emblematic of these conflicting results. % \todo[]{could highlight how influential the LPI has become (recent reports suggest a 73\% ave. pop decline since 1970. This would be massive!}
% With updated data released semi-annually alongside new datasets and new estimates of decline, 
A growing number of high-profile papers have challenged how strong the evidence is for population decline \citep{Dornelas2014,gonzalez2016estimating,wagner2021insect,muller2024weather}, with each paper taking a slightly different analytical approach. For example, using LPI data, \citet{Leung2020} published a mixture model that suggested most populations were not significantly declining, followed by other alternative modeling approaches \citep{Buschke2021,puurtinen2022living} including a recent one suggesting a basic analysis of the dataset should always include three sources of autocorrelation, finding trends that encompassed many previous results \citep{Johnson2024}.

@@

\centerline{\bf Mechanistic forecasting}
\vspace*{2mm}
\begin{minipage}[t]{\linewidth}
    \includegraphics[width=\linewidth]{../figures/forecasting_details}

@@

\noindent % Process-based models are built on explicit mathematical equations to describe (supposedly causal) relationships between environmental drivers and ecological responses. 
% They also often incorporate empirical relationships, particularly when knowledge is incomplete or when some processes are intentionally omitted. Processes are often represented at different nested spatiotemporal scales, depending on the underlying assumptions. 
Model development is the central step of the process-based workflow, typically requiring several years, yet it often remains opaque for anyone who has not worked the model itself. The step of designing the model---translating knowledge and hypotheses into mathematical equations and parameters---is often blurred with the step of model calibration (or tuning), where parameter values are inferred. Models are often treated as an accumulation of multiple submodels, each governing one or several ecological processes (see figure B above). Rather than being fitted as a whole, submodels are calibrated separately against specific subsets of data, and some parameters are simply prescribed (i.e., fixed to a value found in the literature) or tuned to reproduce certain observations or theory. The way models are currently calibrated is likely not a coincidence, but rather a workaround to avoid confronting the full complexity of the model. Calibrating submodels separately allows to avoid the fact that, if the model were fitted as a whole, many parameters would compensate for one another---revealing structural degeneracies and making the model far more difficult to use.  %emw24Jul: nice stuff here.

@@

\centerline{\bf A common workflow to bridge trend estimation and forecasting} 
\vspace*{-3mm}
{\begin{multicols}{2}
\begin{minipage}[t]{\linewidth}
	\vfill
\includegraphics[width=\linewidth]{../figures/iterativeworkflow_details_revised}
\vfill
% \vspace{-0.5cm}\captionsetup{font=footnotesize}\captionof{figure}{Caption}\label{fig:trends}
\vspace*{3mm}
\end{minipage}

@@

\clearpage
\bibliography{../forecastflows.bib}

>>>>

We argue that many current debates and diverging forecasts are driven by the incoherent and disconnected workflows used today in ecology \citep{Loreau2022, Talis2023, Johnson2024}. Research estimating biodiversity trends has become focused on methodological aspects because the current workflow fails to examine the gap between ideal and available data, and rarely tests for predictive accuracy that could scale up to allow forecasting. At the same time, process-based models developed for forecasting often evolve through the addition of new separate layers or components because the current workflow rarely examines the model as a functioning whole and thus ignores major problems (e.g. non-identifiability, discussed below). Thus, it encourages new parts that are often disconnected from the original research aim, its data stream, and the previous scientific insights.

@@

\begin{figure}
	\centering
	\includegraphics{../../figures/figure_modeldataspaces_revised}
	\caption{Different trajectories of current (b-c) and proposed (d) workflows through model-data space during model development, from pre-model/pre-data to post-model/post-data quadrants.
	Panel A serves as a legend, showing where model development should begin (carefully thinking about the hypotheses to test) and when inference should be made, for either explanatory purpose (estimation) or forecasting (prediction).
	Panel B illustrates that, in current trend estimation approaches, most time is spent in the post-model/post-data quadrant, where models are directly fitted to empirical data---often with no prior evaluation of the model.
	Panel C illustrates the typical process-based model development where a lot of time is spent building and calibrating (fitting) separate submodels (represented by the separated ovals) without clear distinction between model building and model calibration and without evaluating the full model structure a priori. Additionally, the specific hypotheses the model is designed to test are rarely clearly stated.
	Panel D illustrates the workflow highlighted in this paper (see main text and Figure 2 for details). Ideally, it begins in the pre-model/pre-data quadrant, where the research question is defined. The next step moves to the post-model/pre-data quadrant, where a model is built---independently of the data. This is where we should spend a good deal of time evaluating model behavior using simulated data. Once we are satisfied with our model, we can move to the post-model/post-data quadrant, where the model is fitted to real data. Then, we perform retrodictive checks to compare model predictions to observations, which likely give us some feedback to refine our model.}
	\label{fig:modeldata}
\end{figure}

@@

Quantitative science relies on a model-based framework to confront hypotheses with data \citep{Chamberlin:1965cd}. In an idealized scientific method, we would formulate a research question and hypotheses, design an experiment accordingly, build a model, collect data, and use this data to inform our model and differentiate between hypotheses. This method underlies much of the recent pre-registration movement, where hypotheses and methodology are defined prior to data collection \citep{Nosek2018}.
But this idealized method often does not apply to the reality of ecological research. Many important questions cannot be addressed through controlled experiments and replications. In such cases, we must rely on existing, heterogeneous datasets alongside uncertain and incomplete theory to provide a large-scale and long-term perspective \citep{Hilborn1997}. Indeed, most macroecological insights have emerged from exploring patterns in these datasets (exploratory data analysis).

This reality should drive researchers to use more robust and coherent methods. But the current workflows combined with the challenges ecologists are facing---both in terms of data complexity and societal needs---instead may lead to persistent problems.
Trend estimation has focused mostly on fitting a model to empirical data (\llabel{quad1}i.e. post-model/post-data, Figure \ref{fig:modeldata}b)---without the checks (\llabel{quad2}post-model/pre-data) and likely feedbacks that often highlight uncertainty and related limitations in the model and/or data (Figure \ref{fig:modeldata}b). For forecasting, researchers have focused on making predictions with increasingly complex mechanistic models (Figure \ref{fig:modeldata}c), frequently obscuring the steps underlying model building and parameterization (\llabel{quad3}and without a clear distinction between pre- and post-data). Researchers often calibrate the different parts of these models separately, and fix some parameter values based on experiments and expert knowledge, to avoid problems when trying to fit the model as a whole.
Addressing these problems while accounting for the realities of working with ecological data requires a more comprehensive workflow.

We argue a workflow that moves \llabel{quad4}along the data-model space in a coherent sequence of steps (Figure \ref{fig:modeldata}d) could reduce many of these problems and thus improve ecological science.
The first step of this workflow is to define an explicit research question and formulate hypotheses (step 1, Figure \ref{fig:workflow}). This involves making clear assumptions about the most influential drivers, within the specific context of our study. This should guide the construction of a narrative model of how we believe the system works, focusing on the mechanisms that could generate the data we observe, including the observational error (\llabel{quad5}pre-model/pre-data, Figure \ref{fig:modeldata}d).  
\llabel{causalinf} This step of carefully formulating hypotheses has gained increasing attention in the ecological literature \citep{Grace2020}. From this narrative, we can then develop a mathematical model---an ensemble of equations that encapsulates our knowledge and is designed to answer our research question (step 2, Figure \ref{fig:workflow}). Generally, starting with a relatively simple model that we could refine later makes understanding the model and how it interacts with empirical data easier \llabel{exampleworkflow1}(see example workflow we provide as a supplementary file). At this stage, prioritizing biologically meaningful parameters is crucial, as it allows us to have a sense of plausible parameter values. This means choosing a model formulation where each parameter corresponds to an interpretable behavior (which sometimes requires considering alternative parameterization).

With a model in place, the next step focuses on testing and understanding it via data simulation (step 3, Figure \ref{fig:workflow}). `Fake' or `test' data are generated directly from the model by fixing parameters \llabel{firstsim}to some reasonable values, which is straightforward if the parameters are interpretable, and from `fake' (simulated) predictor data. 
We then fit our model to this simulated dataset and evaluate its ability to recover the prescribed values (\llabel{quad6}post-model/pre-data, Figure \ref{fig:modeldata}d). At this stage, the focus is on understanding the model, so we may need to spend time thinking about whether each parameter is realistic or not. This is also a step for making sure the model is working as expected (see example workflow). \llabel{repeatsim} Ideally, this data simulation step should be repeated several times with different parameter values, sampled within a reasonable range. \llabel{priorPC} In a Bayesian framework, simulated parameters can be sampled from the prior distribution (allowing us to check the implications of our priors \citealp{Gelman2020}), while in other frameworks they can be sampled from a range of biologically reasonable values (which is very similar to the concept of a prior).

\begin{figure}[h]
	\centering
	\vspace*{-0.5cm}
    \hspace*{-1.5cm}
	\includegraphics{../../figures/figure_worflow_wsteps}
	\caption{The workflow we propose here, which builds from recent advances in workflows \citep[][]{betanworkflow,Gelman2020,Schad2020,grinsztajn2021,vandeschoot2021,Wolkovich2024}, focuses on iterative feedbacks between the research question, model and data. Most research should start at Step 1, with the research question, followed by extensive model building and evaluation using simulated data (Steps 2 and 3) before proceeding to fitting the model on empirical data and examining it through retrodictive checks (Steps 4 and 5). Within this workflow, forecasting is a natural output of the process and not a separate process or one available for only certain modeling approaches.  The integration of observations (the data stream) occurs at Step 4---only after the model has been thoroughly evaluated---and can also highlight opportunities to collect new data and enrich the data stream. This figure is expanded from Figure \ref{fig:modeldata}d.}
	\vspace*{-0cm}
	\label{fig:workflow}
\end{figure}

Once we are confident about our model structure, we can introduce real data as part of an initial model fitting step (step 4, Figure \ref{fig:workflow}). This way, we obtain parameter estimates constrained by observations (\llabel{quad7}post-model/post-data, Figure \ref{fig:modeldata}d). 
These parameter estimates lead to the second data simulation step, this time using our fitted model parameters to generate predictions (step 5, Figure \ref{fig:workflow}). This---which we call a retrodictive check (\citealp{betanworkflow}, also called a \llabel{PPC}posterior predictive check in a Bayesian workflow,  \citealp{Gelman2020})---allows model output to be compared to observations \llabel{suppfig1}(see Supplementary Figure 1). \llabel{howtoPPC} This is a powerful approach, in part because checks should be tailored to assess key features of the data and model output together in regard to our research question, but this also means there is no standard way to do this \citep{Gelman2020}.
It is only once all steps have been completed that we can interpret parameter values with respect to our research question.
The workflow encourages a focus on the full model, where any parameter (such as a trend estimate) must be carefully interpreted alongside others, as all are fundamental components that shape both inference and forecasting.

Within such a workflow, forecasting emerges as a natural outcome. Rather than being a final goal, forecasting only involves jointly modeling new circumstances along with the original data. This workflow could also bridge gaps in certain areas between exploratory analyses and developing research questions. For example, in macroecological studies model building is often informed by patterns in the data. The adoption of this workflow would make the exploratory stage more transparent---as an explicit preliminary step for entering the workflow---and would compel researchers to more clearly develop a research question before extensive model fitting (Figure 1).

A key feature of this workflow is the central role of data simulation, which introduces two feedback loops.
The first feedback arises when we evaluate the model on simulated data.
The failure of the model to recover known parameter values and handle the complexity of the simulated data should prompt reconsidering the model, or even reformulating the research question.
Further, this step might reveal that some parameters are highly non-identifiable (meaning the parameter(s) cannot be uniquely estimated), flagging the need to change the model structure---before incorporating empirical observations. 
The second feedback loop comes from the retrodictive check. 
Discrepancies here may indicate a missing key driver, and suggest the current model is too simplistic. We can refine the model to integrate the missing process(es)---if we can identify them---and return to the start of the workflow. Insights from the retrodictive check can also lead us to introduce additional complexity when simulating fake data, such as phylogenetic structure \llabel{suppfig2}(see Supplementary Figure 1) or observational biases (e.g. unbalanced data). For example, if a researcher realizes their empirical data is geographically biased, this bias should be built into the model and thus into this data simulation step. This iterative evaluation of the model moves beyond a simple reliance on goodness-of-fit metrics. At each iteration, we are able to evaluate the model behavior, both with simulated and real data, taking into account our expert knowledge of the ecological processes. \llabel{broadcontext}These feedback loops are applied currently only in Bayesian frameworks (to our knowledge), but could and should extend across other modeling approaches given their power to improve models and, in turn, scientific understanding.

@@

We argue that an improved workflow that required data simulation and retrodictive checks would lead to larger model advances and a greater recognition of uncertainty---thus highlighting likely consistency in estimates across models---that could better aid policy.  Using the workflow would make it more obvious that what now appear as major discrepancies are shifts in point estimates that are generally all in the same uncertainty space \citep{Johnson2024}---and it would challenge modelers to show major predictive advances, which is not currently part of the process. Explanatory power in most models of observational data is usually very low \citep{low2014rising,moller2002much} and thus tests of models' predictions rarely expected. But the workflow highlights that predictions from the model---what we call retrodictive checks---are part of the process of science, and critical to testing for what may be missing in a model \llabel{suppfig3}(see Supplementary Figure 1).

@@

Applying the workflow to current trend estimates could importantly highlight the best way to improve data collection for more reliable estimates. Returning to the example of a global estimate of trends in vertebrate populations of species over time (see Box) and applying our proposed workflow would mean more efforts to define the goal and question---is it a simple global estimate?
Or a need to also find which species are declining most, including those that may have poor or no data? From there a generative model using simulated data for testing could incorporate many aspects of the populations, and data, that are often only included in `null' or `synthetic data generation' currently \citep{Buschke2021,mcrae2025utility}, but could be built into the models fit to the empirical data. Eventually fitting the empirical data and performing retrodictive checks would likely highlight major missing components of the generative model \llabel{suppfig4}(see Supplementary Figure 1) and, ultimately, this would help inform our global estimates of mean trends \llabel{exampleworkflow2}(see example workflow we provide as a supplement file). For example, certain populations are recovering for very specific reasons (e.g. elephants in regions where the ivory trade drove declines in the past) that perhaps should be modeled. From this model, what data are most critically needed to address the updated aims would become clearer and could drive new data collection \citep{toszogyova2024mathematical}.

@@

Ecological forecasting is a broad field with a diverse range of methods. Across these methods, process-based modeling is often considered the gold standard for forecasting in ecology \citep{Urban2016, Pilowsky2022} and beyond because it is based on mechanistic understanding that should drive accurate out-of-sample predictions.  
This focus on a mechanistic basis, however, generally drives newer models to incorporate greater complexity---and an ever-growing number of parameters. This can make it difficult to increase scientific understanding \citep{Franklin2020} and suggest or model potential policies. 
In principle, increasing model complexity can be beneficial by enhancing mechanistic realism, better representing more complex dynamics, and \llabel{quant2}reducing uncertainty by restricting the effective parameter space. In practice, however, this is not always the outcome. In process-based modeling for climate forecasts, the uncertainty range on the effect of increasing CO$_{2}$ concentration on temperature have remained largely unchanged over decades of increasing model complexity \citep{Zelinka2020}. This has driven calls for more rigorous and transparent calibration processes \citep{balaji2022general}. Similar concerns arise in ecology, where strong disagreements persist about the effect of climate change on future species distributions \citep{Cheaib2012} and ecosystem dynamics \citep{Lovenduski2017} after decades of modeling work.
These uncertainties have large implications beyond ecology, as they influence simulations of biosphere-atmosphere interactions and, ultimately, future climate projections \citep{Bonan2018, simpson2025confronting}.
Some researchers now advocate for simplifying models, to avoid over-parametrization when the data provide little information to constrain some parameters \citep{Wang2017, Harrison2021}. These arguments stress that the strength of process-based models is their interpretability, and thus their parameters should represent biological processes, and the values of these parameters are of interest.
If a model becomes too complex, however, it may become a black box, \llabel{blackbox}where many parameters are effectively hidden and can take on unreasonable values. 
Further, with so many parameters, understanding the sources of uncertainty and how they propagate through the model may become nearly impossible (see Box).
Each additional process and parameter can increase overall uncertainty to the point where model projections lose their usefulness for decision makers \citep{Saltelli2020}.

@@

Applying the workflow to process-based models would help open the black box and thus could identify and fix model problems before forecasting.
Each successive step of model development in the proposed workflow may highlight problems and a path to solutions. Incorporating data simulation would introduce a crucial step between model building and data fitting (also called calibration), ensuring a clear delineation between the two, and help expose potential parameter identifiability issues earlier in the model design. 
Uncovering identifiability issues would likely force researchers to begin with a simpler version of the model, which they could build on iteratively, testing for support---or lack thereof---when adding model complexity. The workflow may also highlight model problems by requiring more explicit model calibration (i.e. data fitting), which is currently hidden within opaque `model building', making it easy to hide non-identifiability. Once found, non-identifiable parts of models could be addressed by reformulating the mathematical structure of certain processes, or finding ways to apply additional constraints (e.g. narrower ranges for certain parameters or developing new hypotheses that target the appropriate level of complexity). The resulting process-based models would likely be simpler and thus more tractable for quantifying parameter uncertainty and propagating uncertainty through to projections.

@@

\centerline{\bf Trend estimation} 
\vspace*{2mm}
\begin{minipage}[t]{\linewidth}
    \includegraphics[width=\linewidth]{../../figures/trendestimation_details}

@@

% Trends - outline current problem
% In the current workflow for estimating trends over time, a new model with a new estimate often leads to a paper (see figure A above) because ecologists spend less time interrogating their models with simulated data, or their model performance fit to empirical data.
In the current workflow for estimating trends over time, a model is typically chosen a priori to fit to data, yielding parameter estimates that are treated as true and often translated directly into published findings and conclusions, without interrogating the model using simulated data and carefully evaluating its performance (see example workflow we provide as a supplementary file). Any such analysis must be treated with caution, because different model variants can yield very different estimates (hence different conclusions), and published results may not prove robust when confronted with future reanalysis of the same data (see Figure A above).
The Living Planet Index (LPI, \url{www.livingplanetindex.org}), which aims to include long-term data on vertebrate populations of species across the globe, is emblematic of these conflicting results. % \todo[]{could highlight how influential the LPI has become (recent reports suggest a 73\% ave. pop decline since 1970. This would be massive!}
% With updated data released semi-annually alongside new datasets and new estimates of decline, 
A growing number of high-profile papers have challenged how strong the evidence is for population decline \citep{Dornelas2014,gonzalez2016estimating,wagner2021insect,muller2024weather}, with each paper taking a slightly different analytical approach. For example, using LPI data, \citet{Leung2020} published a mixture model that suggested most populations were not significantly declining, followed by other alternative modeling approaches \citep{Buschke2021,puurtinen2022living} including a recent one suggesting a basic analysis of the dataset should always include three sources of autocorrelation, finding trends that encompassed many previous results \citep{Johnson2024}.

@@

\centerline{\bf Mechanistic forecasting}
\vspace*{2mm}
\begin{minipage}[t]{\linewidth}
    \includegraphics[width=\linewidth]{../../figures/forecasting_details}

@@

\noindent % Process-based models are built on explicit mathematical equations to describe (supposedly causal) relationships between environmental drivers and ecological responses. 
% They also often incorporate empirical relationships, particularly when knowledge is incomplete or when some processes are intentionally omitted. Processes are often represented at different nested spatiotemporal scales, depending on the underlying assumptions. 
Model development is the central step of the process-based workflow, typically requiring several years, yet it often remains opaque for anyone who has not worked the model itself. The step of designing the model---translating knowledge and hypotheses into mathematical equations and parameters---is often blurred with the step of model calibration (or tuning), where parameter values are inferred. Models are often treated as an accumulation of multiple submodels, each governing one or several ecological processes (see Figure B above). Rather than being fitted as a whole, submodels are calibrated separately against specific subsets of data, and some parameters are simply prescribed (i.e., fixed to a value found in the literature) or tuned to reproduce certain observations or theory. The way models are currently calibrated is likely not a coincidence, but rather a workaround to avoid confronting the full complexity of the model. Calibrating submodels separately allows to avoid the fact that, if the model were fitted as a whole, many parameters would compensate for one another---revealing structural degeneracies and making the model far more difficult to use.  %emw24Jul: nice stuff here.

@@

\centerline{\bf A common workflow to bridge trend estimation and forecasting} 
\vspace*{-3mm}
{\begin{multicols}{2}
\begin{minipage}[t]{\linewidth}
	\vfill
\includegraphics[width=\linewidth]{../../figures/iterativeworkflow_details_revised}
\vfill
% \vspace{-0.5cm}\captionsetup{font=footnotesize}\captionof{figure}{Caption}\label{fig:trends}
\vspace*{3mm}
\end{minipage}

@@

\clearpage
\bibliography{../../forecastflows.bib}
